---
title: "article"
author: "Emanuel Calvo"
date: "December 22, 2015"
output: html_document
runtime: shiny
---

```{r init, message = FALSE, echo=FALSE}
library(shiny)
require(RMySQL)
library(ggplot2)


source('~/git/fts_article/load.R')

# Yeah, only for development environements
all_cons <- dbListConnections(MySQL())
for(con in all_cons) dbDisconnect(con)


# You do not want clear passwords, but is a sandbox.
con <- dbConnect(MySQL(), 
                 user="msandbox", password="msandbox", port=14901 
                 ,dbname="test", socket="/tmp/mysql_sandbox14901.sock", 
                 host="127.0.0.1")


```

This R Markdown document is made interactive using Shiny. Unlike the more traditional workflow of creating static reports, you can now create documents that allow your readers to change the assumptions underlying your analysis and see the results immediately.  To learn more, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).


> You only need to have a MySQL sandbox with the 5.7 version.

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change.  This demonstrates how a standard R plot can be made interactive by wrapping it in the Shiny `renderPlot` function. The `selectInput` and `sliderInput` functions create the input widgets used to drive the plot.

```{r, echo=TRUE}
        ui <- shinyUI(fluidPage(
                mainPanel(textInput("filtering", "FTS Filter:", ""),
                          submitButton("Plot!"),
                          plotOutput("plot")
                          )
        ))
        server <- function(input, output, session) {
                output$plot<-renderPlot({
                        f<-function(x) {as.character(input$filtering)}
                        
                        sql <- paste("SELECT sourceTable, 
        sum(scoreBoolean) as sumBoolean, 
        sum(scoreQE) as sumQE,
        max(scoreBoolean) as maxBoolean,
        max(scoreQE) as maxBoolean,
        avg(scoreBoolean) as avgBoolean, 
        avg(scoreQE) as avgQE,
        avg(lengthContent) avgLength,
        avg(scoreQE) / count(sourceTable) as weight,
        count(sourceTable) as countOccurrences
FROM
(
SELECT \"BC\" as sourceTable, FTS_DOC_ID, match(BC.content) against (\"" ,f(x) ," \" IN BOOLEAN MODE) as scoreBoolean,
       match(BC.content) against (\""  ,f(x), "\" IN NATURAL LANGUAGE MODE WITH QUERY EXPANSION) scoreQE,
       length(content) as lengthContent
FROM
        bookContent BC
WHERE
  match(BC.content) against (\"" ,f(x)," \" IN NATURAL LANGUAGE MODE WITH QUERY EXPANSION)
UNION
SELECT \"BL\" as sourceTable, FTS_DOC_ID, match(BL.content) against (\"",f(x)," \" IN BOOLEAN MODE) as scoreBoolean,
       match(BL.content) against (\"" ,f(x)," \" IN NATURAL LANGUAGE MODE WITH QUERY EXPANSION) scoreQE,
       length(content) as lengthContent
FROM
        bookContentByLine BL
WHERE
  match(BL.content) against (\"", f(x) ,"\" IN NATURAL LANGUAGE MODE WITH QUERY EXPANSION)
) unionTable
GROUP BY sourceTable
;" , sep = "")
                        rs <- dbGetQuery(con, sql)
                        #rs
                        #x<-1:10
                        
                        #pairs(rs$sourceTable + rs$weight, data = rs)
                        pairs(rs[,2:10])
                        # plot(x,f(x))
                }                        
                )              
        }

shinyApp(ui = ui, server = server)
```

Template

```{r}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

```{r, echo=FALSE}
shinyAppDir(
  system.file("examples/06_tabsets", package="shiny"),
  options=list(
    width="100%", height=550
  )
)
```

Note the use of the `height` parameter to determine how much vertical space the embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to prevent the R code within the chunk from rendering in the document alongside the Shiny components.



